% ============================================================
% 00_abstract.tex
% ============================================================

\begin{abstract}

We establish information-theoretic limits on temporal distinguishability
in diffusive and anomalously diffusive systems.

Time is treated not as a primitive variable but as an inferred parameter
of stochastic dynamics, constrained by the distinguishability of
probability distributions under finite spatial resolution, finite
statistics, and measurement noise.

For normal diffusion, Fisher information yields an
estimator-independent lower bound on temporal resolution,
$\Delta t_{\min}(t) \sim t\sqrt{2/(dN)}$,
up to dimensionless constants.
An equivalent local bound follows from hypothesis testing via the
Kullback--Leibler divergence, demonstrating consistency between
estimation-theoretic and testing-based formulations in the
small-$\Delta t$ regime.

In the photon-limited regime, a self-consistent analysis predicts a
nontrivial temporal resolution scaling
$\Delta t_{\min} \propto \Phi^{-1/3}$,
distinct from the standard $1/\sqrt{N}$ noise suppression.
This scaling is derived analytically, verified by numerical simulations,
and is directly testable with existing single-particle tracking and
fluorescence-based measurement techniques.

The framework is extended to anomalous diffusion, revealing qualitatively
distinct temporal regimes governed by the anomalous exponent and the
associated scaling of distinguishability with time.
All results follow from first principles and are falsifiable through
explicit experimental protocols.

Crucially, we show that these limits are not artifacts of particular
estimators or experimental implementations.
Within a fixed observational channel and model class, an explicit
information-theoretic no-go result demonstrates that no inference
procedure can asymptotically surpass the reported scalings without
introducing additional information beyond the spatial statistics
considered here.

The present work provides a minimal operational definition of temporal
resolution in stochastic systems and a general template for analyzing
time distinguishability under fundamental information constraints.

\end{abstract}
