% ============================================================
% 07_falsifiability_protocols.tex
% ============================================================

\section{Falsifiability and Experimental Protocols}
\label{sec:falsifiability}

This work makes quantitative, testable predictions. In this section we formulate
explicit experimental protocols and statistical criteria under which the theory
can be confirmed or falsified within controlled uncertainty bounds. All predictions
are understood to apply within a fixed observational channel and noise model, as
specified in Section~\ref{sec:assumptions_scope}.

\subsection{Core Falsifiable Predictions}

The theory predicts the following universal scaling laws:

\begin{enumerate}
\item \textbf{Statistical regime (CRLB):}
\[
\Delta t_{\min}(t) \sim t \sqrt{\frac{2}{dN}}
\]
for independent samples drawn from a Gaussian propagator.

\item \textbf{Photon-limited regime:}
\[
\Delta t_{\min} \propto \Phi^{-1/3}
\]
independently of $D$, $\alpha$, and dimensionality $d$.

\item \textbf{Anomalous diffusion scaling:}
\[
\Delta t_{\min}(t) \propto
\begin{cases}
t^{1-\alpha}, & \text{resolution-limited (interpretive bound)},\\[4pt]
t^{(2-\alpha)/3}, & \text{photon-limited}.
\end{cases}
\]
\end{enumerate}

Any systematic and statistically significant deviation from these scaling exponents
constitutes a refutation of the corresponding regime.

\subsection{Experimental Platforms}

The predictions can be tested using the following established techniques:

\paragraph{Single-Particle Tracking (SPT).}
Fluorescent tracer particles in aqueous or viscoelastic media, with time inferred
from spatial distributions rather than explicit timestamps.

\paragraph{Fluorescence Correlation Spectroscopy (FCS).}
Temporal resolution inferred indirectly from spatial autocorrelation functions
under controlled photon flux.

\paragraph{Confocal or Light-Sheet Microscopy.}
Direct control of photon flux $\Phi$ and point-spread function $\sigma_0$ within a
well-characterized optical channel.

\subsection{Controlled Experimental Parameters}

Each protocol must explicitly control:

\begin{itemize}
\item Photon flux $\Phi$ (via laser power or exposure time),
\item Effective spatial resolution $\Delta x$ or PSF width $\sigma_0$,
\item Medium properties determining $D$ or $D_\alpha$,
\item Dimensionality $d$ (2D membrane vs.\ 3D bulk),
\item Number of independent trajectories $N$.
\end{itemize}

\subsection{Statistical Test for Scaling Exponents}

Scaling laws are tested via log--log regression. Let
$\Delta t_{\min} \propto X^\beta$ for a control parameter $X$.

\paragraph{Null hypothesis.}
The measured exponent $\hat{\beta}$ differs from the predicted value $\beta_0$.

\paragraph{Test statistic.}
\[
z = \frac{\hat{\beta}-\beta_0}{\mathrm{SE}(\hat{\beta})},
\]
where $\mathrm{SE}(\hat{\beta})$ is the standard error of the regression estimate,
assumed approximately normal in the asymptotic regime.

\paragraph{Rejection criterion.}
The theory is falsified for the tested regime if
$|z| > z_{\alpha/2}$ at confidence level $1-\alpha$
(e.g., $z_{0.025}=1.96$).

\subsection{Photon-Limited Refutation Criterion}

The photon-limited regime is refuted if
\begin{equation}
\frac{d\log \Delta t_{\min}}{d\log \Phi} \neq -\frac{1}{3}
\end{equation}
within statistical uncertainty across at least one order of magnitude in $\Phi$,
under otherwise fixed experimental conditions.

\subsection{Resolution-Limited Refutation Criterion}

The resolution-limited (interpretive) regime is refuted if
\begin{equation}
\Delta t_{\min} \not\propto \Delta x^2
\end{equation}
when $\sigma_0^2 \gg 2dD_\alpha t^\alpha$, after accounting for finite-sample effects
and estimator bias.

\subsection{Anomalous Exponent Validation}

The anomalous exponent $\alpha$ must be independently estimated from the MSD scaling
\[
\langle r^2(t) \rangle \sim t^\alpha.
\]

The framework is falsified if the inferred temporal scaling of $\Delta t_{\min}(t)$
is statistically inconsistent with the independently measured $\alpha$ within
experimental uncertainty.

\subsection{Control Experiments}

The following controls are mandatory:

\begin{itemize}
\item Advection suppression: $\mathrm{Pe} \ll 1$,
\item Motion-blur correction via exposure-time deconvolution,
\item Background-noise subtraction and photobleaching correction,
\item Verification of estimator efficiency via Monte Carlo simulation.
\end{itemize}

\subsection{Summary of Falsification Conditions}

The theory is considered falsified for a given regime if one or more of the following
are robustly observed:

\begin{enumerate}
\item Scaling exponents differ significantly from predictions,
\item Photon-limited $\Phi^{-1/3}$ scaling fails,
\item Resolution-limited $\Delta x^2$ scaling fails,
\item Anomalous exponent $\alpha$ fails cross-consistency,
\item Deviations persist under all control corrections.
\end{enumerate}

These criteria are sufficient to render the framework fully Popper-falsifiable and
experimentally decidable.
